Type-Safe Modular Parsing

### 2017-01-14 paper/sections/PackratParsing.tex
Section 2

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators
are suitable to address those challenges. The algorithm challenges
are important because they rule out various common techniques
used by non-modular code using parsing combinators.
To avoid pitfalls related to those algorithmic challenges,
we propose the following methodology:

@@

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1-2-3$ is parsed as $(1-2)-3$ in the left-recursive
grammar, but after rewrite the result is $1((-2)-3))$. The parse tree
must be transformed to recover its structure.

\paragraph{Longest Match Composition} Another problematic issue
in parser combinator libraries is the need for manually ordering
alternatives in a grammar.
Consider the grammar:
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}

@@

\paragraph{Backtracking} The need for backtracking is also problematic
in a modular setting. For example, consider a grammar that includes
``import'' statements including \inlinecode{import..from}.
Now we want to extend the grammar with an \inlinecode{import..as} case, as shown in the third line of grammar below.

@@

\noindent However, this results in the worst case exponential time complexity in Parsec, because it does not have related optimization.

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately, some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsers use memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, Packrat parsing, in
theory, supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} which implements Packrat parsers.
The library provides a number of parser combinators, including the longeset match alternative combinator.
Below we present an example to illustrate Scala parsers.

@@

\paragraph{Parsing a Simple Arithmetic Language}
Suppose we want to parse a simple language with literals and
additions. The concrete syntax is as follows:

@@

It is straightforward to model the abstract syntax by classes as below. The ASTs support pretty-printing via the \lstinline{print} method.

@@

It is worth mentioning that the grammar is left-recursive in the addition case. Since left-recursion is well supported by Packrat parsers, we do not need extra code for it. Our parser also employs longest match composition by using the combinator \lstinline{|||} in the library. Furthermore, it does not suffer from the backtracking problem, as the memoization technique of Packrat parsing guarantees the efficiency.

>>>>

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators~\cite{Ford2002}
are suitable to address those challenges. The algorithmic challenges
are important because they rule out various common techniques
used by non-modular code using parsing combinators.
To avoid pitfalls related to those algorithmic challenges,
we propose the following methodology:

@@

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1-2-3$ is parsed as $(1-2)-3$ in the left-recursive
grammar, but after rewrite the result is $1((-2)-3))$\bruno{??This is
  not a valid expression}. The parse tree
must be transformed to recover the correct syntactic structure.

\paragraph{Longest Match Composition} Another problematic issue
in parser combinator libraries is the need for manually prioratizing/ordering
alternatives in a grammar.
Consider the grammar:
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}

@@

\paragraph{Backtracking} The need for backtracking can also be problematic
in a modular setting. For example, consider a grammar that includes
``import'' statements including \inlinecode{import..from}.
Now we want to extend the grammar with an \inlinecode{import..as} case, as shown in the third line of grammar below.

@@

\noindent However, this results in the worst case exponential time
complexity in Parsec, because it does not have proive optimizations
for backtracking.

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately, some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsers use memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, Packrat parsing, in
theory, supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} which implements Packrat parsers.
The library provides a number of parser combinators, including the longest match alternative combinator.
%Below we present an example to illustrate Scala parsers.

@@

\paragraph{Parsing a Simple Arithmetic Language}
Suppose we want to parse a simple language with literals and
additions. The concrete syntax is:

@@

It is straightforward to model the abstract syntax by classes. The ASTs support pretty-printing via the \lstinline{print} method.

@@

It is worth mentioning that the grammar is left-recursive in the addition case. Since left-recursion is well supported by Packrat parsers, we do not need extra code for it. Our parser also employs longest match composition by using the combinator \lstinline{|||} in the library. Furthermore, it does not suffer from the backtracking problem, as the memoization technique of Packrat parsing guarantees reasonable efficiency.
